Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-12 analysis/paper/paper.Rmd
updated chunks to estimate abundance and CPUE, added some tables and figures showing results.

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [biomark, mhe]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Under A Bridge, Seattle, Washington, 00000, USA
  - mhe: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

```{r load-libraries}
library(tidyverse)
library(magrittr)
library(lubridate)
library(tidytable)
library(janitor)
library(purrr)
library(here)
library(FSA)

@@

```{r cmr-effort-data}
# cmr_df = read_csv(here("analysis/data/raw_data/deadwater_cmr_effort_20210614.csv")) %>%
cmr_df = readxl::read_excel(here("analysis/data/raw_data/Deadwater_predation_mark_recapture_11.09.21.xlsx")) %>%
  clean_names() %>%
  filter(count > 0) %>%
  # pull out only Northern Pikeminnow
  filter(species == "Northern Pikeminnow") %>%
  # mutate(across(date,
  #               mdy)) %>%
  mutate(across(date,
                ymd)) %>%
  mutate(across(c(start_time, end_time),
                ~ ymd_hms(paste(year(date), month(date), day(date), hour(.), minute(.), second(.))))) %>%
  # split into sampling periods
  mutate(event_name = if_else(month(date) > 8,
                              paste("Fall", year(date), sep = "_"),
                              paste("Spring", year(date), sep = "_")))

effort_df = cmr_df %>%
  select(event_name,
         site,
         date,
         crew,
         method,
         equipment,
         start_time:total_effort,
         total_day_effort) %>%
  distinct() %>%
  group_by(event_name,
           date) %>%
  summarise(across(total_effort,
                   sum),
            .groups = "drop")

cpue_df = effort_df %>%
  # add how many pikeminnow were caught
  left_join(cmr_df %>%
              filter(capture_type != "Recapture") %>%
              group_by(event_name, species, date) %>%
              summarise(n_fish = n(),
                        .groups = "drop")) %>%
  mutate(cpue = n_fish / total_effort)

@@

LP_mod = cmr_df %>%
  filter(species == "Northern Pikeminnow") %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  mutate(mr_type = if_else(cap_period == "Mark" & fish_status == "Live",
                           "M",
                           if_else(cap_period == "Recapture",
                                   if_else(capture_type == "Mark",
                                           "C",
                                           if_else(capture_type == "Recapture",
                                                   "R",
                                                   NA_character_)),
                                   NA_character_))) %>%
  filter(!is.na(mr_type)) %>%
  group_by(event_name, species) %>%
  summarise(M = sum(mr_type == "M"),
            n = sum(mr_type %in% c("C", "R")),
            m = sum(mr_type == 'R'),
            .groups = "drop") %>%
  nest(data = -c(event_name, species)) %>%
  crossing(model = c("Petersen",
                     "Chapman")) %>%
  mutate(mr_mod = map2(data,
                       model,
                      .f = function(x, y) {
                        with(x,
                             mrClosed(M = M,
                                      n = n,
                                      m = m,
                                      method = y))
                      }),
         N_summ = map(mr_mod,
                      .f = summary,
                      incl.SE = T),
         CI = map(mr_mod,
                  .f = confint)) %>%
  mutate(N = map_dbl(N_summ,
                     .f = function(x) x[1]),
         SE = map_dbl(N_summ,
                      .f = function(x) x[2]),
         N.lci = map_dbl(CI,
                         .f = function(x) x[1]),
         N.uci = map_dbl(CI,
                         .f = function(x) x[2])) %>%
  select(-N_summ, -CI)
```

```{r schnabel-estimator, eval = F}
# using Schnabel model
N_mods = cmr_df %>%
  filter(species == "Northern Pikeminnow") %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species, date, fish_status, capture_type) %>%
  summarise(n_fish = n(),
            .groups = "drop") %>%
  group_by(event_name, species, date) %>%
  summarise(n = sum(n_fish),
            m = sum(n_fish[capture_type == "Recapture"]),
            R = sum(n_fish[fish_status == "Live"]),
            dead_fish = sum(n_fish[fish_status == "Dead"]),
            .groups = "drop") %>%
  group_by(event_name, species) %>%
  nest() %>%
  mutate(mod = map(data,
                   .f = function(x) {
                     with(x,
                          mrClosed(n = n,
                                   m = m,
                                   R = R,
                                   method = "Schnabel",
                                   chapman.mod = TRUE))
                   }),
         N = map_dbl(mod,
                     .f = summary),
         CI = map(mod,
                  .f = confint))

N_mods %>%
  select(species, N, CI) %>%
  mutate(N.lci = map_dbl(CI,
                         .f = function(x) x[1]),
         N.uci = map_dbl(CI,
                         .f = function(x) x[2])) %>%
  select(-CI)
```

@@

```{r shoup-temp-data}
temp_df = read_tsv("../data/raw_data/USGS_13307000_Salmon_River_Nr_Shoup.txt",
                   comment = "#") %>%
  rename(discharge_cfs = `47768_00060`,
         discharge_cfs_cd = `47768_00060_cd`,
         gage_ht_ft = `47769_00065`,
         gage_ht_ft_cd = `47769_00065_cd`,
         water_temp_c = `47773_00010`,
         water_temp_c_cd = `47773_00010_cd`)

@@

>>>>

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [biomark, mhe]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Under A Bridge, Seattle, Washington, 00000, USA
  - mhe: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

```{r load-libraries}
library(tidyverse)
library(magrittr)
library(lubridate)
library(tidytable)
library(janitor)
library(purrr)
library(here)
library(FSA)
library(knitr)
library(kableExtra)

# format NA values in tables
options(knitr.kable.NA = '-')

# set default theme for figures
theme_set(theme_bw())

@@

```{r cmr-effort-data}
raw_df = readxl::read_excel(here("analysis/data/raw_data/Deadwater_predation_mark_recapture_11.09.21.xlsx")) %>%
  clean_names() %>%
  mutate(across(date,
                ymd)) %>%
  mutate(across(c(start_time, end_time),
                ~ ymd_hms(paste(year(date), month(date), day(date), hour(.), minute(.), second(.))))) %>%
  # split into sampling periods
  mutate(event_name = if_else(month(date) > 8,
                              paste("Fall", year(date), sep = "_"),
                              paste("Spring", year(date), sep = "_")))

# focus on pikeminnow
cmr_df = raw_df %>%
  filter(count > 0) %>%
  # pull out only Northern Pikeminnow
  filter(species == "Northern Pikeminnow") %>%
  mutate(mark_release = if_else(fish_status == "Live" &
                                  capture_type == "Mark",
                                T, F),
         recap = if_else(capture_type == "Recapture",
                         T, F))

effort_df = raw_df %>%
  select(event_name,
         site,
         date,
         crew,
         crew_count,
         method,
         equipment,
         start_time:total_effort,
         total_day_effort) %>%
  distinct() %>%
  mutate(across(total_effort,
                ~ . * crew_count)) %>%
  group_by(event_name,
           date) %>%
  summarise(across(total_effort,
                   sum),
            .groups = "drop")

cpue_df = effort_df %>%
  # add how many pikeminnow were caught
  left_join(cmr_df %>%
              group_by(event_name, species, date) %>%
              summarise(n_fish = n(),
                        .groups = "drop")) %>%
  mutate(cpue = n_fish / total_effort)

@@

LP_mod = cmr_df %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species) %>%
  summarise(M = sum(cap_period == "Mark" & fish_status == "Live", na.rm = T),
            n = sum(cap_period == "Recapture"),
            m = sum(cap_period == "Recapture" & capture_type == "Recapture", na.rm = T),
            .groups = "drop") %>%
  nest(data = -c(event_name, species)) %>%
  crossing(model = c("Petersen",
                     "Chapman")) %>%
  mutate(mr_model = map2(data,
                       model,
                      .f = function(x, y) {
                        with(x,
                             mrClosed(M = M,
                                      n = n,
                                      m = m,
                                      method = y))
                      }),
         N_summ = map(mr_model,
                      .f = summary,
                      incl.SE = T),
         CI = map(mr_model,
                  .f = confint)) %>%
  mutate(N = map_dbl(N_summ,
                     .f = function(x) x[1]),
         SE = map_dbl(N_summ,
                      .f = function(x) x[2]),
         Lci = map_dbl(CI,
                         .f = function(x) x[1]),
         Uci = map_dbl(CI,
                         .f = function(x) x[2])) %>%
  select(-N_summ, -CI)

```{r schnabel-estimator, eval = F}
# using Schnabel model
N_mods = cmr_df %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species, date) %>%
  summarise(n = n(),
            m = sum(capture_type == "Recapture", na.rm = T),
            u = sum(capture_type == "Mark" | is.na(capture_type)),
            R = sum(fish_status == "Live", na.rm = T),
            .groups = "drop") %>%
  group_by(event_name, species) %>%
  nest() %>%
  crossing(model = c("Schnabel",
                     "SchumacherEschmeyer")) %>%
  mutate(mr_model = map2(data,
                         model,
                         .f = function(x, y) {
                           with(x,
                                mrClosed(n = n,
                                         m = m,
                                         R = R,
                                         method = y,
                                         chapman.mod = TRUE))
                         }),
         N = map_dbl(mr_model,
                     .f = summary),
         CI = map(mr_model,
                  .f = confint),
         Lci = map_dbl(CI,
                       .f = function(x) x[1]), 
         Uci = map_dbl(CI,
                       .f = function(x) x[2])) %>%
  ungroup() %>%
  select(-CI)

@@

```{r shoup-temp-data}
temp_df = read_tsv(here("analysis/data/raw_data/USGS_13307000_Salmon_River_Nr_Shoup.txt"),
                   comment = "#") %>%
  rename(discharge_cfs = `47768_00060`,
         discharge_cfs_cd = `47768_00060_cd`,
         gage_ht_ft = `47769_00065`,
         gage_ht_ft_cd = `47769_00065_cd`,
         water_temp_c = `47773_00010`,
         water_temp_c_cd = `47773_00010_cd`)

@@

```{r abund-est}
all_est = LP_mod %>%
  select(event_name, species, model, N:Uci) %>%
  bind_rows(N_mods %>%
             select(event_name, species, model, N:Uci)) %>%
  mutate(range = Uci - Lci,
         prop_range = range / N) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  mutate(model = fct_recode(model,
                            "Schumacher-Eschmeyer" = "SchumacherEschmeyer")) %>%
  arrange(event_name, model)
```

```{r abund-tab}
all_est %>%
  select(-c(species, range:prop_range)) %>%
  rename(`Sampling Event` = event_name,
         Estimator = model) %>%
  kbl(booktabs = T,
      format.args = list(big.mark = ",")) %>%
  kable_styling()
```

```{r abund-fig, fig.cap = "Estimates of abundance of northern pikeminnow using different estimators."}
all_est %>%
  mutate(across(c(N, Lci, Uci),
                 ~ . / 1000)) %>%
  ggplot(aes(x = model,
             y = N,
             color = model)) +
  geom_errorbar(aes(ymin = Lci,
                    ymax = Uci),
                width = 0.1) +
  geom_point(size = 4) +
  facet_wrap(~ event_name) +
  scale_color_brewer(palette = "Set1") +
  coord_cartesian(ylim = c(0, 100)) +
  labs(x = "Estimator",
       y = "Pikeminnow Abundance (x 1,000)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))

```{r cpue-fig, fig.cap = "Total catch per unit effort across entire sampling event."}
cpue_df %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  group_by(species, event_name) %>%
  summarise(across(c(total_effort, n_fish),
                   sum),
            .groups = "drop") %>%
  mutate(across(total_effort,
                ~ . / 60)) %>%
  mutate(cpue = n_fish / total_effort) %>%
  ggplot(aes(x = event_name,
             y = cpue,
             fill = event_name)) +
  geom_col() +
  scale_fill_brewer(palette = "Set1",
                    name = "Event") +
  labs(x = "Event",
       y = "CPUE (Pikeminnow caught per hour)") +
  theme(legend.position = "none")
